---
title: "Project 3: Count Regression (Bike Sharing System)"
author: "David O. Agboola"
#date: "March 23, 2019"
header-includes:
    - \usepackage{setspace}\doublespacing
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{center}
\bf\large{PART I}
\end{center}

## Question i.

The mean-variance relationship refers to the property that for count distributions, the variation is often calculated using the mean. In other words, the variation changes depending on the value of the mean.

For the Poisson, the mean and variance are the same. For the Negative Binomial, the variance is a multiple larger than 1 times the mean.

## Question ii.

Overdispersion means there is more dispersion (or loosely speaking, variability) in the data than the model can accommodate. For instance, there is overdispersion if the variance is larger than the mean in the Poisson regression model since the Poisson has equal mean and variance. This is a concern in count regression models because when it happens, the model underestimates the variablity in the data. This leads to a very large test statistics since the denominators will be very small. Thus, we have an inflated Type I error, leading to false declarations of significance. 


## Question iii.

The overdispersed Poisson distribution assumes a different distribution for the rate parameter $\lambda$. The distribution assumed typically has two parameters that account for the shape and scale of the rate parameter. This is similar to Bayesian technique where a parameter of interest has a prior distribution assigned to it for better estimation. Thus making the variance a multiple larger than $1$ times the mean. Hence, overdispersed Poisson distribution has more variability than the Poisson distribution. 

## Question iv. 

When a count regression model have extremely large positive residuals, it means that the response variable (or $y-$values) are not fitted closely. This would suggest that the model is a poor fit.

## Question v.

The coefficient ($\beta$) within a count regression model is interpreted by first exponentiating it. So, $e^\beta$ represents the expected multiplicative change in the mean counts (for a unit increase of the predictor). This interpretation would involve "odds ratios" since we usually compare populations.  

## Question vi.

For a Hurdle model, the count regression coefficient $\beta$ (which represents the logistic component of the model) gives odds ratios for being in zero-population; while the count regression coefficient $\gamma$ (which represents the truncated count component of the model) gives changes in mean count for those in the non-zero-population. This is possible because the Hurdle model describes individuals from two populations.

For a Zero-Inflated model, the interpretation is opposite that of Hurdle model because Zero-Inflated model may not describe individuals from two populations. This is because one regression coefficient $\beta$ (which represents the logistic component of the model) gives odds ratios for being in zero-population;  while the count regression coefficient $\gamma$ gives changes in mean count for any number including those in the zero-population.


\newpage
\begin{center}
\bf\large{PART II}
\end{center}

\begin{center}
\bf\large{Purpose of the Analysis}
\end{center}

Bike sharing systems are getting popular and becoming a new generation of traditional bike rentals where the whole process from membership, rental and return has become automatic – easier than before. Through these systems, a user can easily rent a bike from a position and return the bike at another position without stress or paperwork. Currently, there are about over 500 bike-sharing programs around the world which is composed of over 500 thousand bicycles. There are recent research interests in these systems due to their important role in traffic, environmental and health issues.
It is of interest to know what variables influence the average number of bikes being rented daily. For this project, we plan to consider the variables: weather situation, holiday, feeling temperature, humidity, and wind-speed; so as to know what combination of these variables can be utilized to predict the average number of bikes to be rented on a given day.

\begin{center}
\bf\large{Description of the Data}
\end{center}

The data-set “Bike-Sharing-Data-set” was collected by the Laboratory of Artificial Intelligence and Decision Support (LIAAD), University of Porto which contains the daily count of rental bikes for the years 2011 and 2012 from Capital Bikeshare system, Washington D.C., USA (http://capitalbikeshare.com/system-data); the sample size is 731.  The aggregated data with the corresponding holiday, weather and seasonal information was also collected by the Laboratory of Artificial Intelligence and Decision Support (LIAAD), University of Porto. The data-set was downloaded from UCI Machine Learning Repository https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset. Weather information were extracted from http://www.freemeteo.com.

Our response variable is count with predictors – weather situation, holiday, feeling temperature, humidity, and wind-speed.
The data type of weather situation is integer where 1 represents "cool weather" – clear and few clouds;  2 represents "not so cool weather" – fog, mist, and partly cloudy; 3 represents "not cool weather" – snow, ice pallets, rain, and thunderstorm. The data type of holiday is integer (extracted from http://dchr.dc.gov/page/holiday-schedule) where 0 represents "no holiday" and 1 represents "holiday". The data type of feeling temperature, humidity, and wind-speed is number measured in Celsius, percentage, and m/sec respectively. And lastly, the data type of count is integer representing count of total rental bikes.


\begin{center}
\bf\large{Proposed Analysis}
\end{center}

For exploratory data analysis, we would first check for missing values or duplicated data, then a histogram and boxplot of the response variable will be utilized to see how the variable is distributed. For descriptive data analysis, the five-number summary of the response variable will be observed. Then, using the matrix scatterplot with correlation we check if feeling temperature correlate with humidity, if feeling temperature correlate with wind-speed, and if humidity correlate with wind-speed. A matrix scatterplot with correlation of wind-speed, humidity, feeling temperature and count will also be considered. This will suggest what predictor variables to drop to avoid redundancy and what predictors to try for predicting the response.

The Negative Binomial Regression model:
\[
\ln\mu_i=\beta_0+\beta_1\cdot weathersit + \beta_2 \cdot holiday + \beta_3 \cdot TEMPfel + \beta_4 \cdot Humid + \beta_5 \cdot Windspeed.
\]
would be used because our response variable is a count data and zero-truncated. Also, Negative Binomial is the most appropriate because it accommodates the extra variation parameter that exceeds that allowed by the Poisson Count Regression model. Negative Binomial Regression model would help fit an appropriate equation for predicting the average number of bikes to be rented daily given the predictors.

Inferential analysis involves regressing the appropriate predictors with the response and checking if the Negative Binomial Regression model is a good fit. We check by observing the range of values of the deviance residuals if it falls within the standard normal range of -3 to 3, observing the quotient of the deviance residuals and degrees of freedom if the value is close to 1, and plotting the deviance residuals against the fitted response to observe the overall variation in the model – this is good for visuals. From the deviance residuals and visuals, we observe if there are anomalies and explain them. So, if the model suggests a good fit, then we can trust the result it gives. Using Wald test, we find out what predictors are statistically significant and use their estimates (the $\beta's$) for specific interpretations.


\begin{center} 
\bf\large{Analysis Results}
\end{center}

Using the histogram and boxplot, we observed that the response variable follows a roughly symmetric-shaped distribution with no observable outlier. Since it is a count data, it is also zero-truncated. We checked and observed no missing values or duplicated data.

The fitted Negative Binomial Regression model is:
\[
\ln\mu_i=8.149-0.137\cdot weathersit -0.199 \cdot holiday + 0.039 \cdot TEMPfel -0.005 \cdot Humid -0.015 \cdot Windspeed.
\]

The table below shows the exponentiated value of the estimates with the p-values of the parameters excluding the intercept: 

Parameters   | Exp(Estimates)|  P-values
-------------| ------------- | -------------
Weathersit   | $0.872$       | $$6.03 \times 10^{-5}$$
Holiday      | $0.820$       | $0.0191$
TEMPfel      | $1.040$       | $$ < 2 \times 10^{-16}$$
Humid        | $0.995$       | $0.00012$
Windspeed    | $0.985$       | $$1.47 \times 10^{-7}$$.

We observed that all the predictors are statistically significant at $95\%$ confidence level.

Thus, for an increase in bad weather situation, expect the average number of bikes rented to reduce by 13% (or change by a multiple of approximately 0.87). For holidays, expect the average number of bikes rented to reduce by 18% (or change by a multiple of approximately 0.82 times when there is no holiday). For an increase in feeling temperature by $1^0$ Celsius, expect the average number of bikes rented to increase by about 4% (or change by a multiple of approximately 1.04). For an increase in humidity by $1\%$, expect the average number of bikes rented to reduce by about $1\%$ (or change by a multiple of approximately 0.99). And lastly, for an increase in wind-speed by 1 meter per seconds, expect the average number of bikes rented to reduce by about 2% (or change by a multiple of approximately 0.98).

Hence, we can predict with 95% confidence level the average number of bikes to be rented on a given day using the predictors – weather situation, holiday, feeling temperature, humidity, and wind-speed.

It was observed that the quotient of the deviance residual to degrees of freedom is close to 1 (precisely $1.03$); deviance residual ranged from $-6.99$ to $2.85$ and by using visuals for the deviance residuals we observed that one observation was unusually low ("cnt"=22) which resulted in the very low extreme at $-6.99.$ Further exploratory analysis indicated the date in which this record was made (October 29, 2012) which was when Washington D.C. was observing a state of emergency and government shutdown due to possible effects of Hurricane Sandy. This greatly reduced the number of bikes rented that day. Hence, the data is an anomaly and was not removed. The next lowest residual is $-3.79$ which is not too far away from $-3$. Since the observation makes sense, our result suggests a good fit, so we can trust our results.


\begin{center} 
\bf\large{Conclusions}
\end{center}

From the results above, we can conclude that weather situation, holiday, feeling temperature, humidity, and wind-speed are key factors/variables that influence the average number of bikes rented daily. As the weather situation deteriorates, the average number of bikes rented reduces which makes sense. It is interesting to observe that more bikes are rented when there is no holiday than when there is holiday. It will be interesting to know why this is the case and what class of people rents bikes more. For every increase in wind-speed and humidity, we observe a reduction in the average number of bikes rented. And when temperature increases, the average number of bikes rented is also likely to increase.
With the combination of these variables, we can predict with some certainty the average number of bikes to be rented on a given day.


\newpage
\begin{center} 
\bf\large{APPENDIX}
\end{center}

```{r, echo=T,warning=FALSE,message=FALSE,tidy=TRUE,echo=FALSE}
setwd("C:\\Users\\agboo\\OneDrive - University of Northern Colorado\\Spring 2019\\SRM 611\\Project")
data <- read.csv("day.csv",header=TRUE,sep=",")
#head(data)
hist(data$cnt,main="Histogram of number of bikes rented",xlab = "total number of bikes")
boxplot(data$cnt,main="Boxplot for number of bikes rented",ylab="total number of bikes");summary(data$temp)
plot(density(data$cnt),main="Density plot for bikes rented")
summary(data$cnt) #count summary
sum(is.na(data));sum(is.null(data)) #check if we have missing values
sum(duplicated(data))
data$TEMPref<-data$temp * 41 #converted normalized temperature in Celsius
data$TEMPfel<-data$atemp*50 #cOnverted normalized feeling temperature in Celsius
data$Humid<-data$hum*100 #converted normalized humidity
data$WINDSPEED<- data$windspeed*67 #converted normalized windspeed
data$weathersit1<-ifelse(data$weathersit==1|data$weathersit==2,0,1) #recode weather situation with dummy variables
data$weathersit2<-ifelse(data$weathersit==2,1,0)

library(s20x)
plot(data$Humid,data$TEMPfel,xlab = "Humidity",ylab = "Feeling Temperature")
cor(data$Humid,data$TEMPfel)
plot(data$WINDSPEED,data$TEMPfel,xlab = "Wind-speed",ylab = "Feeling Temperature")
cor(data$WINDSPEED,data$TEMPfel)
plot(data$Humid,data$WINDSPEED,xlab = "Humidity",ylab = "Wind-speed")
cor(data$Humid,data$WINDSPEED)
DAT<-cbind(data$cnt,data$WINDSPEED,data$Humid,data$TEMPfel)
colnames(DAT)<-c("Count","Windspeed","Humidity","Tempfel")
pairs20x(DAT) #check correlation among variables
# NEGATIVE BINOMIAL MODEL #
library("MASS")
negativeBinomialReg = glm.nb(cnt~as.factor(weathersit)+holiday+TEMPfel+Humid+WINDSPEED,data=data)
summary(negativeBinomialReg)
exp(summary(negativeBinomialReg)$coefficients[,1])
# DEVIANCE RESIDUALS #
nbDevResid = summary(negativeBinomialReg)$deviance.resid
summary(nbDevResid)
library("ggplot2")
qplot(negativeBinomialReg$y,nbDevResid)
summary(data$cnt)
sum(is.na(data));sum(is.null(data)) #check if we have missing values
negativeBinomialReg = glm.nb(cnt~weathersit+holiday+TEMPfel+Humid+WINDSPEED,data=data)
summary(negativeBinomialReg)
exp(summary(negativeBinomialReg)$coefficients[,1])
nbDevResid = summary(negativeBinomialReg)$deviance.resid
summary(nbDevResid)
```
